---
title: "Predicting ATP Tennis Match Outcomes: Logistic Regression, Elo, and Hybrid Models"
author: "Himanshu Muley (A1895211)"
date: "`r Sys.Date()`"
output:
  pdf_document:
    toc: true
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

# Setup and Libraries

```{r libraries}
library(tidyverse)
library(dplyr)
library(elo)
library(glmnet)
library(yardstick)
library(inspectdf)
```


---

# Data Loading and Cleaning

We load all annual ATP match files from 2000 to 2019 and combine them into
a single dataset. This section also defines the outcome variable and all
shared features used by every model downstream.

```{r load-data}
# Load each year
file_years <- 2000:2019
tennis_list <- lapply(file_years, function(y) {
  read_csv(paste0("./data/atp_matches_", y, ".csv"), show_col_types = FALSE)
})
combined <- bind_rows(tennis_list)
```

```{r select-and-clean}
t_data <- combined %>%
  select(
    winner_name, loser_name,
    winner_rank, loser_rank,
    winner_rank_points, loser_rank_points,
    surface, score, tourney_date, tourney_level
  ) %>%
  mutate(year = as.integer(substr(tourney_date, 1, 4)))

# Remove rows with any missing values
t_data <- drop_na(t_data)

# data types of the columns
t_data <- t_data %>%
  mutate(
    winner_rank = as.integer(winner_rank),
    loser_rank  = as.integer(loser_rank),
    winner_rank_points = as.integer(winner_rank_points),
    loser_rank_points = as.integer(loser_rank_points),
    surface   = as.factor(surface),
    tourney_level = as.factor(tourney_level)
  )
```

## Outcome Definition — Important Note

We define `player_i` and `player_j` by **randomly assigning** which player
takes each role per match. This avoids the tautology present in a naive
implementation where `player_i` is always the winner (which would make the
outcome perfectly predictable from any winner-derived feature).

Under this scheme:

- `outcome = 1` if `player_i` wins the match
- `outcome = 0` if `player_j` wins the match

The ranking-point difference `rp_diff = rp_i - rp_j` can now be positive
or negative, giving the logistic model genuine signal to learn from rather
than a trivially positive number.

```{r outcome-definition}
set.seed(42)  # for reproducibility

t_data <- t_data %>%
  mutate(
    # Randomly decide which player is i and which is j
    swap = sample(c(TRUE, FALSE), n(), replace = TRUE),

    player_i  = if_else(swap, loser_name,         winner_name),
    player_j  = if_else(swap, winner_name,         loser_name),
    rp_i      = if_else(swap, loser_rank_points,   winner_rank_points),
    rp_j      = if_else(swap, winner_rank_points,  loser_rank_points),
    rank_i    = if_else(swap, loser_rank,           winner_rank),
    rank_j    = if_else(swap, winner_rank,           loser_rank),

    # outcome = 1 if player_i won
    outcome     = if_else(swap, 0L, 1L),

    # ranking point difference from player_i's perspective
    rp_diff = as.integer(rp_i - rp_j)
  ) %>%
  mutate(
    outcome  = as.factor(outcome),
    surface = as.factor(surface),
    tourney_level = as.factor(tourney_level)
  ) %>%
  rename(tour_lvl = tourney_level)

inspect_na(t_data)
```

##Plots

```{r plot-matches-per-year}
ggplot(t_data, aes(x = year)) +
  geom_bar(fill = "#4B9CD3") +
  theme_minimal(base_size = 14) +
  labs(
    title = "Number of ATP Matches per Year (2000–2019)",
    x = "Year",
    y = "Match Count"
  )
```

```{r plot-surface}
ggplot(t_data, aes(x = surface)) +
  geom_bar(fill = "#6A5ACD") +
  theme_minimal(base_size = 14) +
  labs(
    title = "Match Count by Surface Type",
    x     = "Surface",
    y     = "Match Count"
  )
```
 

## Train / Test Split

Matches from 2000–2018 are used for training. The 2019 season is held out
as the test set, consistent with prior literature and avoiding COVID-19
disruptions in later years.

```{r train-test-split}
train <- t_data %>% filter(year >= 2000, year <= 2018)
test  <- t_data %>% filter(year == 2019)


dim(train) #dimentions of the training dataset
```
```{r}
dim(test) #dimentions of the testing dataset
```

---

# Shared Evaluation Functions

We define all evaluation metrics once and reuse them across every model.

```{r eval-functions}
# Accuracy at 0.5 threshold
accuracy_from_prob <- function(y_true, p, threshold = 0.5) {
  pred_class <- ifelse(p >= threshold, 1, 0)
  mean(pred_class == y_true)
}

# Log-loss (with numerical clipping for stability)
log_loss <- function(y, p) {
  eps <- 1e-15
  p   <- pmin(pmax(p, eps), 1 - eps)
  -mean(y * log(p) + (1 - y) * log(1 - p))
}

# Brier score
brier_score <- function(y, p) {
  mean((p - y)^2)
}

# Calibration table (equal-width bins)
calibration_table <- function(y, p, bins = 10) {
  df      <- data.frame(y = y, p = p)
  df$bin  <- cut(df$p, breaks = seq(0, 1, length.out = bins + 1),
                 include.lowest = TRUE)
  df %>%
    group_by(bin) %>%
    summarise(
      mean_pred   = mean(p),
      mean_actual = mean(y),
      count       = n(),
      .groups     = "drop"
    )
}

# Calibration plot function
plot_calibration <- function(calib_df, title, colour = "blue") {
  ggplot(calib_df, aes(x = mean_pred, y = mean_actual)) +
    geom_point(size = 3, colour = colour) +
    geom_abline(slope = 1, intercept = 0,
                linetype = "dashed", colour = "red") +
    labs(
      title = title,
      x     = "Mean Predicted Probability",
      y     = "Empirical Win Rate"
    ) +
    theme_minimal()
}


```

---

# Part A — Logistic Regression Models

## Baseline Logistic Regression

The baseline model uses only the ranking-point difference (`rp_diff`) as a
predictor, with no intercept. This mirrors the specification in Klaassen
and Magnus (2003) and ensures a 50% predicted probability when both
players have equal ranking points.

```{r logit-baseline-fit}
logit_base <- glm(
  outcome ~ 0 + rp_diff,
  data   = train,
  family = binomial(link = "logit")
)

summary(logit_base)


```

```{r logit-baseline-test}
# Align factor levels between train and test
test <- test %>%
  mutate(
    surface   = factor(surface,   levels = levels(train$surface)),
    tour_lvl  = factor(tour_lvl,  levels = levels(train$tour_lvl))
  )

# Predict on test set
test$prob_base <- predict(logit_base, newdata = test, type = "response")
test$pred_base <- factor(ifelse(test$prob_base > 0.5, 1, 0), levels = c(0, 1))
test$outcome   <- factor(test$outcome, levels = c(0, 1))

# Metrics
y_base <- as.integer(as.character(test$outcome))
p_base <- test$prob_base

 accuracy_from_prob(y_base, p_base)

```
```{r}
 log_loss(y_base, p_base)

```

```{r}
brier_score(y_base, p_base)
```

```{r logit-baseline-roc}
test_roc_base <- test %>%
  mutate(outcome = factor(outcome, levels = c(0, 1)))
```


```{r}
roc_base <- roc_curve(test_roc_base, truth = outcome, prob_base,
                      event_level = "second")
autoplot(roc_base) +
  labs(
    title = "ROC Curve – Baseline Logistic Regression (2019 Test Set)",
    x     = "1 - Specificity (False Positive Rate)",
    y     = "Sensitivity (True Positive Rate)"
  ) +
  theme_minimal(base_size = 14)

roc_auc(test_roc_base, truth = outcome, prob_base, event_level = "second")
```

## Extended Logistic Regression

The extended model adds surface type and tournament level as additional
predictors, motivated by evidence that playing conditions influence ATP
match outcomes (Kovalchik, 2016).

```{r}
logit_ext <- glm(outcome ~ 0 + rp_diff + surface + tour_lvl,data = train,
family = binomial(link = "logit")
)

summary(logit_ext)

```

```{r}
anova(logit_ext)
```


```{r logit-extended-test}
test$prob_ext1 <- predict(logit_ext, newdata = test, type = "response")
test$pred_ext1 <- factor(ifelse(test$prob_ext1 > 0.5, 1, 0), levels = c(0, 1))

y_ext1 <- as.integer(as.character(test$outcome))
p_ext1 <- test$prob_ext1


 accuracy_from_prob(y_ext1, p_ext1) 


 
```
```{r}
log_loss(y_ext1, p_ext1)
```

```{r}
 brier_score(y_ext1, p_ext1)
```

```{r logit-extended-roc}
roc_ext1 <- roc_curve(test, truth = outcome, prob_ext1,
                      event_level = "second")
autoplot(roc_ext1) +
  labs(
    title = "ROC Curve – Extended Logistic Regression (2019 Test Set)",
    x     = "1 - Specificity (False Positive Rate)",
    y     = "Sensitivity (True Positive Rate)"
  ) +
  theme_minimal(base_size = 14)

roc_auc(test, truth = outcome, prob_ext1, event_level = "second")
```

---

# Part A — Elo Rating Models

For the Elo models, the outcome must be an integer (0/1) rather than a
factor. We also use `player_i` and `player_j` as defined in the shared
data prep above, ensuring consistency with the logistic models.

## Baseline Elo Model

All players begin with a rating of 1500 at the start of 2000. Ratings are
updated sequentially through each match using K = 32. After training, the
final 2018 ratings are transferred to the test set.

```{r elo-train-prep}
train_elo <- train %>%
  mutate(outcome_int = as.integer(as.character(outcome))) %>%
  arrange(year)
```

```{r elo-baseline-train}
elo_base_run <- elo.run(
  outcome_int ~ player_i + player_j,
  data    = train_elo,
  k       = 32,
  verbose = FALSE
)

elo_base_df <- as.data.frame(elo_base_run)

# Store final 2018 ratings
ratings_2018 <- final.elos(elo_base_run)
final_ratings <- data.frame(
  player = names(ratings_2018),
  rating = as.numeric(ratings_2018)
)
```

```{r elo-baseline-test-prep}
test_elo <- test %>%
  mutate(outcome_int = as.integer(as.character(outcome))) %>%
  arrange(year)

# Build starting Elo vector for test — default 1500, overwrite with 2018 ratings
all_players        <- sort(unique(c(train_elo$player_i, train_elo$player_j,
                                    test_elo$player_i,  test_elo$player_j)))
all_players        <- all_players[!is.na(all_players)]
elos_test          <- rep(1500, length(all_players))
names(elos_test)   <- all_players
common             <- intersect(names(ratings_2018), names(elos_test))
elos_test[common]  <- ratings_2018[common]
```

```{r elo-baseline-test}
elo_base_test_run <- elo.run(
  outcome_int ~ player_i + player_j,
  data         = test_elo,
  k            = 0,               # no rating updates on test set
  initial.elos = elos_test,
  verbose      = FALSE
)

elo_base_test_df <- as.data.frame(elo_base_test_run)

test_elo <- test_elo %>%
  mutate(
    elo_i_before   = elo_base_test_df$elo.A,
    elo_j_before   = elo_base_test_df$elo.B,
    elo_prob_base  = elo_base_test_df$p.A,
    elo_diff       = elo_i_before - elo_j_before
  )

y_elo  <- test_elo$outcome_int
p_elo  <- test_elo$elo_prob_base


 accuracy_from_prob(y_elo, p_elo)
 log_loss(y_elo, p_elo)
 
```
```{r}
brier_score(y_elo, p_elo)
```

```{r elo-baseline-calibration}
# Note: we use quantile-based bins so every point represents the same
# number of matches — this gives a fuller, more informative calibration plot.
calib_elo_base <- calibration_table(y_elo, p_elo, bins = 10)
cat("Calibration table — number of non-empty bins:",
    sum(!is.na(calib_elo_base$mean_pred)), "\n")
calib_elo_base

plot_calibration(calib_elo_base,
                 "Calibration Curve – Baseline Elo (2019)",
                 colour = "blue")
```
A calibration curve checks whether your predicted probabilities are trustworthy.
Each dot is a group of matches. You take all matches where the model predicted 
around the same probability (say 0.20–0.30), group them into a bin, and then 
check: out of those matches, how many did player ii i actually win?
If the model is perfect: a dot at predicted 0.25 should have an actual win rate
of 0.25 too. Every dot would sit on the red dashed line.

What the chart shows:

The dot around predicted 0.25 has an actual win rate of ~0.37 — the model said 
"25% chance" but the player actually won 37% of the time. The model is too 
pessimistic here.The dots around 0.60–0.75 sit below the line — the model said 
"70% chance" but the player only won ~60% of the time. The model is too 
confident here.



## Extended Elo Model (Surface-Specific K)

The K-factor is varied by surface to reflect differing levels of match
volatility across surfaces, following Kovalchik (2020).

```{r elo-extended-train}
train_elo_ext <- train_elo %>%
  mutate(
    k_surface = case_when(
      surface == "Clay"   ~ 36,
      surface == "Grass"  ~ 24,
      surface == "Hard"   ~ 32,
      surface == "Carpet" ~ 28,
      TRUE                ~ 32
    )
  )

elo_ext_run <- elo.run(
  outcome_int ~ player_i + player_j,
  data    = train_elo_ext,
  k       = train_elo_ext$k_surface,
  verbose = FALSE
)

ratings_2018_ext        <- final.elos(elo_ext_run)
elos_test_ext           <- rep(1500, length(all_players))
names(elos_test_ext)    <- all_players
common_ext              <- intersect(names(ratings_2018_ext), names(elos_test_ext))
elos_test_ext[common_ext] <- ratings_2018_ext[common_ext]
```

```{r elo-extended-test}
elo_ext_test_run <- elo.run(
  outcome_int ~ player_i + player_j,
  data         = test_elo,
  k            = 0,
  initial.elos = elos_test_ext,
  verbose      = FALSE
)

elo_ext_test_df <- as.data.frame(elo_ext_test_run)

test_elo <- test_elo %>%
  mutate(elo_prob_ext = elo_ext_test_df$p.A)

y_elo_ext <- test_elo$outcome_int
p_elo_ext <- test_elo$elo_prob_ext


 accuracy_from_prob(y_elo_ext, p_elo_ext)
 log_loss(y_elo_ext, p_elo_ext)

```
```{r}
 brier_score(y_elo_ext, p_elo_ext)
```

```{r elo-extended-calibration}
calib_elo_ext <- calibration_table(y_elo_ext, p_elo_ext, bins = 10)
plot_calibration(calib_elo_ext,
                 "Calibration Curve – Extended Elo, Surface-K (2019)",
                 colour = "darkgreen")
```

---

# Part B — Elo-Logistic Hybrid Model

## Motivation

The logistic models in Part A used ATP ranking-point difference as the
sole predictor. While interpretable, this feature is static — it does not
capture player form over time. The Elo rating system, by contrast, updates
dynamically after each match and reflects recent performance.

In Part B, we use the **Elo ratings trained on 2000–2018** as features
inside a logistic regression model. Specifically, for each 2019 test match
we have:

- elo_diff — the difference in pre-match Elo ratings (player_i minus player_j)
- elo_prob_base — the Elo-derived win probability for player_i

This hybrid approach combines the dynamic, history-aware nature of Elo
with the statistical interpretability of logistic regression, and allows
us to compare it directly against the Part A models using the same
evaluation metrics.+

## Hybrid Model: Elo Difference as Predictor

```{r hybrid-prep}
# Attach Elo features to training set using training Elo values
train_elo_df <- as.data.frame(elo_base_run)

train_hybrid <- train_elo %>%
  mutate(
    elo_i_before  = train_elo_df$elo.A,
    elo_j_before  = train_elo_df$elo.B,
    elo_prob_train = train_elo_df$p.A,
    elo_diff      = elo_i_before - elo_j_before
  )
```

```{r hybrid-fit}
# Baseline hybrid: elo_diff only, no intercept (mirrors logistic baseline)
hybrid_base <- glm(
  outcome_int ~ 0 + elo_diff,
  data   = train_hybrid,
  family = binomial(link = "logit")
)

summary(hybrid_base)
```

```{r hybrid-test}
# Generate predicted probabilities on the 2019 test set
test_elo$prob_hybrid_base <- predict(
  hybrid_base,
  newdata = test_elo,
  type    = "response"
)

# Extract actual outcomes and predicted probabilities
y_hyb  <- test_elo$outcome_int
p_hyb  <- test_elo$prob_hybrid_base


 accuracy_from_prob(y_hyb, p_hyb)
 log_loss(y_hyb, p_hyb)
 
```
```{r}
brier_score(y_hyb, p_hyb)
```

```{r hybrid-roc}
# Convert outcome to factor for ROC analysis

test_elo_roc <- test_elo %>%
  mutate(outcome_fac = factor(outcome_int, levels = c(0, 1)))

# Compute ROC curve for hybrid baseline
roc_hybrid <- roc_curve(test_elo_roc,
                        truth        = outcome_fac,
                        prob_hybrid_base,
                        event_level  = "second")
autoplot(roc_hybrid) +
  labs(
    title = "ROC Curve – Hybrid Model (Elo diff in Logistic Regression)",
    x     = "1 - Specificity (False Positive Rate)",
    y     = "Sensitivity (True Positive Rate)"
  ) +
  theme_minimal(base_size = 14)


```
```{r}
roc_auc(test_elo_roc,
        truth       = outcome_fac,
        prob_hybrid_base,
        event_level = "second")

```

```{r hybrid-calibration}
calib_hybrid <- calibration_table(y_hyb, p_hyb, bins = 10)
plot_calibration(calib_hybrid,
                 "Calibration Curve – Hybrid Model (2019)",
                 colour = "purple")
```

## Extended Hybrid: Elo Difference + Surface + Tournament Level

```{r hybrid-ext-fit}
# Extended Hybrid (Elo diff + surface + tournament level)

hybrid_ext <- glm(
  outcome_int ~ 0 + elo_diff + surface + tour_lvl,
  data   = train_hybrid,
  family = binomial(link = "logit")
)

summary(hybrid_ext)

```
```{r}
anova(hybrid_ext)
```

```{r hybrid-ext-test}
# Predict on test set and evaluate

test_elo$prob_hybrid_ext <- predict(
  hybrid_ext,
  newdata = test_elo,
  type    = "response"
)

y_hyb_ext <- test_elo$outcome_int
p_hyb_ext <- test_elo$prob_hybrid_ext


accuracy_from_prob(y_hyb_ext, p_hyb_ext)
log_loss(y_hyb_ext, p_hyb_ext)
 
```

```{r}
 brier_score(y_hyb_ext, p_hyb_ext)
```


```{r hybrid-ext-calibration}
calib_hybrid_ext <- calibration_table(y_hyb_ext, p_hyb_ext, bins = 10)
plot_calibration(calib_hybrid_ext,
                 "Calibration Curve – Extended Hybrid Model (2019)",
                 colour = "darkred")
```

## Interaction Model: Elo Difference × Surface

Adding an interaction term `elo_diff:surface` allows the model to learn
that Elo difference may be a stronger predictor on some surfaces than
others. For example, clay court matches tend to favour the higher-rated
player more consistently, while grass court results are more volatile
due to the serve-dominant nature of the surface (Kovalchik, 2020).

```{r hybrid-interact-fit}
 #Interaction Model (Elo diff × surface)
# Allows the effect of Elo difference to vary by surface
hybrid_interact <- glm(
  outcome_int ~ 0 + elo_diff + elo_diff:surface,
  data   = train_hybrid,
  family = binomial(link = "logit")
)

summary(hybrid_interact)
```

```{r hybrid-interact-test}
test_elo$prob_hybrid_interact <- predict(
  hybrid_interact,
  newdata = test_elo,
  type    = "response"
)

y_hyb_int <- test_elo$outcome_int
p_hyb_int <- test_elo$prob_hybrid_interact

accuracy_from_prob(y_hyb_int, p_hyb_int)
log_loss(y_hyb_int, p_hyb_int)
brier_score(y_hyb_int, p_hyb_int)
```

```{r hybrid-interact-calibration}
calib_hybrid_int <- calibration_table(y_hyb_int, p_hyb_int, bins = 10)
plot_calibration(calib_hybrid_int,
                 "Calibration Curve – Interaction Hybrid (2019)",
                 colour = "orange")
```

## Combined Model: Ranking Points + Elo Difference

This model uses both `rp_diff` (ATP ranking-point difference) and
`elo_diff` (Elo rating difference) as predictors. The rationale is that
these two features capture different signals: ranking points reflect
official ATP standings over the past 12 months, while Elo reflects
cumulative match-by-match form that updates dynamically. A player
returning from injury may have high ranking points but low Elo, or a
rising player may have the opposite. Following Hvattum and Arntzen (2010),
who used Elo as a logistic regression covariate, we extend their approach
by including both predictors simultaneously.

```{r combined-prep}
# Attach rp_diff to the training hybrid data
# (rp_diff is already in the original train data, we need it in train_hybrid)
train_hybrid$rp_diff <- train_elo$rp_diff
```

```{r combined-fit}
hybrid_combined <- glm(
  outcome_int ~ 0 + rp_diff + elo_diff,
  data   = train_hybrid,
  family = binomial(link = "logit")
)

summary(hybrid_combined)
```

```{r combined-test}
test_elo$prob_combined <- predict(
  hybrid_combined,
  newdata = test_elo,
  type    = "response"
)

y_comb <- test_elo$outcome_int
p_comb <- test_elo$prob_combined+

accuracy_from_prob(y_comb, p_comb)
log_loss(y_comb, p_comb)
brier_score(y_comb, p_comb)
```

```{r combined-roc}
test_elo_roc <- test_elo %>%
  mutate(outcome_fac = factor(outcome_int, levels = c(0, 1)))

roc_combined <- roc_curve(test_elo_roc,
                          truth       = outcome_fac,
                          prob_combined,
                          event_level = "second")
autoplot(roc_combined) +
  labs(
    title = "ROC Curve – Combined Model: rp_diff + elo_diff (2019)",
    x     = "1 - Specificity (False Positive Rate)",
    y     = "Sensitivity (True Positive Rate)"
  ) +
  theme_minimal(base_size = 14)

roc_auc(test_elo_roc,
        truth       = outcome_fac,
        prob_combined,
        event_level = "second")
```

```{r combined-calibration}
calib_combined <- calibration_table(y_comb, p_comb, bins = 10)
plot_calibration(calib_combined,
                 "Calibration Curve – Combined rp_diff + elo_diff (2019)",
                 colour = "darkblue")
```

## Full Extended Combined Model

The final model incorporates all available features: both player-strength
signals (`rp_diff` and `elo_diff`), contextual variables (`surface` and
`tour_lvl`), and interaction terms (`elo_diff:surface`) to allow the
effect of Elo to vary across playing surfaces. This represents the most
comprehensive specification in the project.

```{r full-combined-fit}
hybrid_full <- glm(
  outcome_int ~ 0 + rp_diff + elo_diff + surface + tour_lvl + elo_diff:surface,
  data   = train_hybrid,
  family = binomial(link = "logit")
)

summary(hybrid_full)
```

```{r full-combined-anova}
anova(hybrid_full)
```

```{r full-combined-test}
test_elo$prob_full <- predict(
  hybrid_full,
  newdata = test_elo,
  type    = "response"
)

y_full <- test_elo$outcome_int
p_full <- test_elo$prob_full

accuracy_from_prob(y_full, p_full)
log_loss(y_full, p_full)
brier_score(y_full, p_full)
```

```{r full-combined-roc}
test_elo_roc <- test_elo %>%
  mutate(outcome_fac = factor(outcome_int, levels = c(0, 1)))

roc_full <- roc_curve(test_elo_roc,
                      truth       = outcome_fac,
                      prob_full,
                      event_level = "second")
autoplot(roc_full) +
  labs(
    title = "ROC Curve – Full Extended Combined Model (2019)",
    x     = "1 - Specificity (False Positive Rate)",
    y     = "Sensitivity (True Positive Rate)"
  ) +
  theme_minimal(base_size = 14)

roc_auc(test_elo_roc,
        truth       = outcome_fac,
        prob_full,
        event_level = "second")
```

```{r full-combined-calibration}
calib_full <- calibration_table(y_full, p_full, bins = 10)
plot_calibration(calib_full,
                 "Calibration Curve – Full Extended Combined (2019)",
                 colour = "darkgreen")
```

## Intercept Models 

 allowing a nonzero intercept to improve
calibration. All previous models used `~ 0 + ...` which forces a 50%
prediction when all predictors are zero. Removing this constraint lets
the model learn a baseline win probability, which may improve how
well the predicted probabilities match actual outcomes.

We test this on the three key Part B models where calibration was poorest.

### Hybrid Baseline with Intercept

```{r hybrid-intercept-fit}
hybrid_base_int <- glm(
  outcome_int ~ elo_diff,
  data   = train_hybrid,
  family = binomial(link = "logit")
)

summary(hybrid_base_int)
```

```{r hybrid-intercept-test}
test_elo$prob_hyb_base_int <- predict(
  hybrid_base_int,
  newdata = test_elo,
  type    = "response"
)

y_hbi <- test_elo$outcome_int
p_hbi <- test_elo$prob_hyb_base_int

accuracy_from_prob(y_hbi, p_hbi)
log_loss(y_hbi, p_hbi)
brier_score(y_hbi, p_hbi)
```

```{r hybrid-intercept-calibration}
calib_hbi <- calibration_table(y_hbi, p_hbi, bins = 10)
plot_calibration(calib_hbi,
                 "Calibration Curve – Hybrid Baseline with Intercept (2019)",
                 colour = "steelblue")
```

### Combined Model with Intercept

```{r combined-intercept-fit}
hybrid_combined_int <- glm(
  outcome_int ~ rp_diff + elo_diff,
  data   = train_hybrid,
  family = binomial(link = "logit")
)

summary(hybrid_combined_int)
```

```{r combined-intercept-test}
test_elo$prob_comb_int <- predict(
  hybrid_combined_int,
  newdata = test_elo,
  type    = "response"
)

y_ci <- test_elo$outcome_int
p_ci <- test_elo$prob_comb_int

accuracy_from_prob(y_ci, p_ci)
log_loss(y_ci, p_ci)
brier_score(y_ci, p_ci)
```

```{r combined-intercept-calibration}
calib_ci <- calibration_table(y_ci, p_ci, bins = 10)
plot_calibration(calib_ci,
                 "Calibration Curve – Combined with Intercept (2019)",
                 colour = "navy")
```

### Full Model with Intercept

```{r full-intercept-fit}
hybrid_full_int <- glm(
  outcome_int ~ rp_diff + elo_diff + surface + tour_lvl + elo_diff:surface,
  data   = train_hybrid,
  family = binomial(link = "logit")
)

summary(hybrid_full_int)
```

```{r full-intercept-anova}
anova(hybrid_full_int)
```

```{r full-intercept-test}
test_elo$prob_full_int <- predict(
  hybrid_full_int,
  newdata = test_elo,
  type    = "response"
)

y_fi <- test_elo$outcome_int
p_fi <- test_elo$prob_full_int

accuracy_from_prob(y_fi, p_fi)
log_loss(y_fi, p_fi)
brier_score(y_fi, p_fi)
```

```{r full-intercept-calibration}
calib_fi <- calibration_table(y_fi, p_fi, bins = 10)
plot_calibration(calib_fi,
                 "Calibration Curve – Full Model with Intercept (2019)",
                 colour = "darkgreen")
```

---

# Visualisations for Report

```{r overlaid-roc, fig.cap="ROC curves for the three core modelling approaches evaluated on the 2019 test set. All models show similar discriminative ability, with AUC values between 0.665 and 0.669."}
# Overlaid ROC curves for key models
roc_data_base <- roc_curve(
  test %>% mutate(outcome = factor(outcome, levels = c(0, 1))),
  truth = outcome, prob_base, event_level = "second"
) %>% mutate(Model = "Logistic baseline")

roc_data_elo <- roc_curve(
  test_elo %>% mutate(outcome_fac = factor(outcome_int, levels = c(0, 1))),
  truth = outcome_fac, elo_prob_base, event_level = "second"
) %>% mutate(Model = "Elo baseline")

roc_data_hyb <- roc_curve(
  test_elo %>% mutate(outcome_fac = factor(outcome_int, levels = c(0, 1))),
  truth = outcome_fac, prob_hybrid_base, event_level = "second"
) %>% mutate(Model = "Hybrid baseline")

roc_combined_plot <- bind_rows(roc_data_base, roc_data_elo, roc_data_hyb)

ggplot(roc_combined_plot, aes(x = 1 - specificity, y = sensitivity, 
                               colour = Model, linetype = Model)) +
  geom_line(size = 1, alpha = 0.8) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "grey50") +
  labs(
    title = "ROC Curves — Logistic vs Elo vs Hybrid (2019 Test Set)",
    x = "1 - Specificity (False Positive Rate)",
    y = "Sensitivity (True Positive Rate)"
  ) +
  scale_colour_manual(values = c("Logistic baseline" = "#4B9CD3",
                                  "Elo baseline" = "#E8A838",
                                  "Hybrid baseline" = "#8B5CF6")) +
  scale_linetype_manual(values = c("Logistic baseline" = "solid",
                                    "Elo baseline" = "dashed",
                                    "Hybrid baseline" = "dotted")) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "bottom")

```

```{r logloss-barchart, fig.cap="Log-loss comparison across all 12 models. Lower values indicate better probabilistic calibration. The logistic baseline achieves the lowest log-loss (0.655), while hybrid and combined models show substantially worse calibration."}
# Bar chart of log-loss across all models
logloss_data <- tibble(
  Model = c(
    "Logistic\nbaseline", "Logistic\nextended",
    "Elo\nbaseline", "Elo\nsurface-K",
    "Hybrid\nbaseline", "Hybrid\nextended",
    "Hybrid\ninteraction", "Combined\nrp+elo",
    "Full\nextended",
    "Hybrid\n+intercept", "Combined\n+intercept", "Full\n+intercept"
  ),
  Log_Loss = c(
    log_loss(y_base, p_base), log_loss(y_ext1, p_ext1),
    log_loss(y_elo, p_elo), log_loss(y_elo_ext, p_elo_ext),
    log_loss(y_hyb, p_hyb), log_loss(y_hyb_ext, p_hyb_ext),
    log_loss(y_hyb_int, p_hyb_int), log_loss(y_comb, p_comb),
    log_loss(y_full, p_full),
    log_loss(y_hbi, p_hbi), log_loss(y_ci, p_ci), log_loss(y_fi, p_fi)
  ),
  Group = c(
    "Part A", "Part A", "Part A", "Part A",
    "Part B", "Part B", "Part B", "Part B", "Part B",
    "Part B (intercept)", "Part B (intercept)", "Part B (intercept)"
  )
)

logloss_data$Model <- factor(logloss_data$Model, levels = logloss_data$Model)

ggplot(logloss_data, aes(x = Model, y = Log_Loss, fill = Group)) +
  geom_col(width = 0.7) +
  geom_hline(yintercept = 0.6931, linetype = "dashed", colour = "red", size = 0.5) +
  annotate("text", x = 12, y = 0.698, label = "No-skill (0.693)", colour = "red", size = 3, hjust = 1) +
  labs(
    title = "Log-Loss Comparison Across All Models",
    y = "Log-Loss (lower = better)",
    x = NULL
  ) +
  scale_fill_manual(values = c("Part A" = "#4B9CD3",
                                "Part B" = "#8B5CF6",
                                "Part B (intercept)" = "#E8A838")) +
  theme_minimal(base_size = 12) +
  theme(axis.text.x = element_text(size = 8),
        legend.position = "bottom")
```

```{r calibration-comparison, fig.cap="Calibration curves for the logistic baseline (left) and hybrid baseline (right). Points on the dashed diagonal indicate perfect calibration. The logistic baseline tracks the diagonal more closely, while the hybrid model shows greater deviation."}
# Side-by-side calibration curves
calib_logistic <- calibration_table(y_base, p_base, bins = 10) %>%
  mutate(Model = "Logistic baseline")
calib_hybrid_compare <- calibration_table(y_hyb, p_hyb, bins = 10) %>%
  mutate(Model = "Hybrid baseline")

calib_both <- bind_rows(calib_logistic, calib_hybrid_compare)

ggplot(calib_both, aes(x = mean_pred, y = mean_actual, colour = Model)) +
  geom_point(size = 3) +
  geom_abline(slope = 1, intercept = 0, linetype = "dashed", colour = "red") +
  facet_wrap(~Model) +
  labs(
    title = "Calibration Curves — Logistic Baseline vs Hybrid Baseline",
    x = "Mean Predicted Probability",
    y = "Empirical Win Rate"
  ) +
  scale_colour_manual(values = c("Logistic baseline" = "#4B9CD3",
                                  "Hybrid baseline" = "#8B5CF6")) +
  theme_minimal(base_size = 14) +
  theme(legend.position = "none")
```

---

# Model Comparison

```{r comparison-table}
results <- tibble(
  Model = c(
    "Logistic (baseline)",
    "Logistic (extended)",
    "Elo (baseline)",
    "Elo (surface-K)",
    "Hybrid: Elo diff only",
    "Hybrid: Elo diff + surface + tour",
    "Hybrid: Elo diff × surface interaction",
    "Combined: rp_diff + elo_diff",
    "Full: rp + elo + surface + tour + interact",
    "Hybrid baseline + intercept",
    "Combined + intercept",
    "Full + intercept"
  ),
  Accuracy = c(
    accuracy_from_prob(y_base,    p_base),
    accuracy_from_prob(y_ext1,    p_ext1),
    accuracy_from_prob(y_elo,     p_elo),
    accuracy_from_prob(y_elo_ext, p_elo_ext),
    accuracy_from_prob(y_hyb,     p_hyb),
    accuracy_from_prob(y_hyb_ext, p_hyb_ext),
    accuracy_from_prob(y_hyb_int, p_hyb_int),
    accuracy_from_prob(y_comb,    p_comb),
    accuracy_from_prob(y_full,    p_full),
    accuracy_from_prob(y_hbi,     p_hbi),
    accuracy_from_prob(y_ci,      p_ci),
    accuracy_from_prob(y_fi,      p_fi)
  ),
  Log_Loss = c(
    log_loss(y_base,    p_base),
    log_loss(y_ext1,    p_ext1),
    log_loss(y_elo,     p_elo),
    log_loss(y_elo_ext, p_elo_ext),
    log_loss(y_hyb,     p_hyb),
    log_loss(y_hyb_ext, p_hyb_ext),
    log_loss(y_hyb_int, p_hyb_int),
    log_loss(y_comb,    p_comb),
    log_loss(y_full,    p_full),
    log_loss(y_hbi,     p_hbi),
    log_loss(y_ci,      p_ci),
    log_loss(y_fi,      p_fi)
  ),
  Brier_Score = c(
    brier_score(y_base,    p_base),
    brier_score(y_ext1,    p_ext1),
    brier_score(y_elo,     p_elo),
    brier_score(y_elo_ext, p_elo_ext),
    brier_score(y_hyb,     p_hyb),
    brier_score(y_hyb_ext, p_hyb_ext),
    brier_score(y_hyb_int, p_hyb_int),
    brier_score(y_comb,    p_comb),
    brier_score(y_full,    p_full),
    brier_score(y_hbi,     p_hbi),
    brier_score(y_ci,      p_ci),
    brier_score(y_fi,      p_fi)
  )
)

results <- results %>%
  mutate(across(where(is.numeric), ~round(.x, 4)))

knitr::kable(
  results,
  caption = "Predictive performance of all models on 2019 ATP test set.
             Lower log-loss and Brier score indicate better probabilistic accuracy.",
  col.names = c("Model", "Accuracy", "Log-Loss", "Brier Score")
)
```

---

# Discussion

The key improvement in this combined analysis over Part A is the correction
of the outcome variable definition. In the original implementation, player_i
was always assigned as the winner, which created a data leakage issue: the
ranking-point difference predictor (derived from winner/loser points) was
trivially positive for all outcome = 1 cases, producing artificially
perfect accuracy and AUC. By randomly
assigning player roles, the models now face a genuine prediction task.


